# Experimental design


## Sample size by simulation

From: Study Design, 2hr - Gordana


<br>

Using pilot data, find values of linear model parameters. In this experiment, I would like to see if there is an effect of Petal length on Sepal length. I have pilot data with 15 observations.


```{r}

data(iris)

str(iris)

```

```{r}

table(iris$Species)

```



Get some pilot data (or get different data from Gordana's)

```{r}

# rand_setosa <- sample(1:50, size = 15, replace = F)

# pilot <- iris[rand_setosa, ]

##  Gordana's data

pilot <- iris[1:15, c(1, 3)]

```


```{r}

pilot_mod <- lm(Sepal.Length ~ Petal.Length, data = pilot)

summary(pilot_mod)

beta_0 <- coef(pilot_mod)[1] # estimated intercept (from pilot) 

sd <- summary(pilot_mod)$sigma # estimated variability

my_range <- range(pilot$Petal.Length) # range of petal length

nsim <- 500 # number of simulated datasets

```


Then specify a meaningful effect size (here it’s the slope of Petal length)


```{r}

# ecologically meaningful slope (same scale as data)

beta_1 <- 1

```


```{r}

N <- 20 # desired sample size

sim_dat = data.frame(Sepal.Length = NA, 
                     Petal.Length = seq(my_range[1], 
                                        my_range[2], 
                                        length = N) )

pval = rep(NA, nsim)

for (i in 1:nsim) {
  mean_y <- beta_0 + beta_1 * sim_dat$Petal.Length
  sim_dat$Sepal.Length <- rnorm(N, mean = mean_y, sd = sd)
  m <- lm(Sepal.Length ~ Petal.Length, data = sim_dat)
  pval[i] <- coef(summary(m))["Petal.Length", "Pr(>|t|)"]
  } # cycle through all N values

sum(pval < 0.05) / nsim

```

