# ggplot


## Vignettes

[Extending ggplot2](https://cran.r-project.org/web/packages/ggplot2/vignettes/extending-ggplot2.html)

[Aesthetic specifications](https://cran.r-project.org/web/packages/ggplot2/vignettes/ggplot2-specs.html)

[Cheat sheet](https://rstudio.com/resources/cheatsheets/)


## Confidence bands

https://janhove.github.io/reporting/2017/05/12/visualising-models-2



##  Confidence bands: fitted models

Part of this comes from Gordana's code for calculating sample size by simulation

```{r}

library(dplyr)
library(emmeans)
library(ggplot2)
library(ggpubr)


data(iris, package = "datasets")

str(iris)

```



```{r}

table(iris$Species)

```



```{r}

iris_versi <- filter(iris, Species == "versicolor")

summary(iris_versi)

```



```{r}

iris_versi$Species <- factor(iris_versi$Species)

nsamp <- 12

iris_versi_samp <- sample(1:50, size = nsamp, replace = F)

pilot <- iris_versi[iris_versi_samp, ]


##  Gordana's simulation code

pilot_mod <- lm(Sepal.Length ~ Petal.Length, data = pilot)

summary(pilot_mod)

```



```{r}

confint(pilot_mod)  

```



```{r}

fit_gg <- ggplot(pilot, aes(x = Petal.Length, 
                            y = Sepal.Length)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Fitted value with 95 % confidence 
          interval (conditional mean)") + 
  theme(plot.title = element_text(size = 8, face = "bold"))

print(fit_gg)

```


Check type of band (confidence or prediction) on above ggplot


```{r}

# range of petal length
my_range <- range(pilot$Petal.Length) 

X_pred <- seq(my_range[1], my_range[2], length = 12)

yhat_conf <- data.frame(
  predict(pilot_mod, interval = "confidence", 
          newdata = data.frame(Petal.Length = X_pred)))

names(yhat_conf) <- c("yhat_conf_fit", "yhat_conf_lwr", "yhat_conf_upr")


yhat_pred <- data.frame(
  predict(pilot_mod, interval = "prediction", 
          newdata = data.frame(Petal.Length = X_pred)))

names(yhat_pred) <- c("yhat_pred_fit", "yhat_pred_lwr", "yhat_pred_upr")


yhat_all <- data.frame(Petal.Length = X_pred, yhat_conf, yhat_pred[ , -1])

names(yhat_all)[names(yhat_all) %in% "yhat_conf_fit"] <- "Sepal.Length"

##  Add to sample data

pilot$X_pred <- X_pred

pilot <- data.frame(pilot, yhat_conf)

pilot <- data.frame(pilot, 
                    yhat_pred_lwr = yhat_pred$yhat_pred_lwr,
                    yhat_pred_upr = yhat_pred$yhat_pred_upr)

##  Plot fitted values and intervals from "predict" function

fit_conf <- ggplot(pilot, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point() +
  geom_smooth(
    data = yhat_all, aes(ymin = yhat_conf_lwr, 
                         ymax = yhat_conf_upr), 
    stat = "identity") +
  ggtitle("Fitted value with 95 % confidence 
          interval (from 'predict')")+ 
  theme(plot.title = element_text(size = 8, face = "bold"))


fit_pred <- ggplot(pilot, aes(x = Petal.Length, y = Sepal.Length)) + 
  geom_point() +
  geom_smooth(data = yhat_all, aes(ymin = yhat_pred_lwr, 
                                   ymax = yhat_pred_upr), 
              stat = "identity") +
  ggtitle("Fitted value with 95 % prediction 
          interval (from 'predict')")+ 
  theme(plot.title = element_text(size = 8, face = "bold"))

ggarrange(fit_gg, fit_conf, fit_pred, ncol = 3)

```


<br>


##  Histogram

###  Histogram: one vector of values

```{r}

##  Histogram of one vector of values in ggplot2

##  Generate a vector of values

s2 <- numeric(5000)
for(i in 1:5000) {
  dd <- rnorm(5, 10, 1)
  s2[i] <- var(dd)
  rm(dd)
}

summary(s2)

```


Data in numeric vector

```{r}

##  Data in numeric vector

ggplot() + 
  aes(s2) + 
  geom_histogram(binwidth = 0.5)

```


Data in single-column data frame - Y axis is frequency

```{r}

##  Data in single-column data frame

s2.df <- data.frame(var = s2)

##  Y axis is frequency

ggplot(s2.df, aes(x = var)) +
  geom_histogram(breaks = 0:16*0.5, col = "black", 
                 fill = "grey70")


```


Data in single-column data frame - Y axis is density


```{r}

##  Y axis is density

ggplot(s2.df, aes(x = var, y = ..density..)) +
  geom_histogram(breaks = 0:16*0.5, col = "black", 
                 fill = "grey70")


```

For comparison, histogram in lattice

```{r}

##  Lattice

histogram(s2, breaks = 0:16*0.5, right = F, main = "Lattice")

range(s2)
```


##  Density plot

```{r}

##  Density plot

ggplot(s2.df, aes(x = var)) +
  geom_density()


```

For comparison, density plot in base graphics and in lattice


```{r}

##  Density plot - base graphics

plot(density(s2), main = "Base graphics")


```


```{r}

##  Density plot - lattice

densityplot(s2, plot.points = F, main = "Lattice")


```


### Histogram: vertical axis with percentage scale


```{r message = F, warning = F}

data(Orthodont)

#summary(Orthodont)

ggplot(Orthodont, aes(distance)) + 
  geom_histogram(aes(y = stat(count / sum(count)) * 100),
                   breaks = seq(from = 16, to = 32, by = 2),
                   colour = "black", fill = "grey") +
  scale_x_continuous(breaks = seq(from = 16, to = 32, by = 2)) +
  labs(x = "Distance", y = "Percentage")

```


##  Plot structure

```
ggplot_build(p_obj)$data
```


## Label legend

```
guides(fill = guide_legend(title = "Gender"))
```


## Plot title 

```
theme(plot.title = element_text(size = 11))
```


## Three-category colours

```
scale_fill_manual(values =  c("#f8766d", "#00bfc4", "#b79f00")
```


## Set breaks for scale

```
scale_y_continuous(breaks = seq(0, 12, by = 2)) 
```


## Display table to 2 d.p.

```
mutate_if(is.numeric, ~ round(., 1)) %>%
kable()
```


## Redefine factor levels for an individual plot

```
nt_youth_all_indiv %>%
  mutate(Race.nat = factor(
    Race.nat, 
    levels = c("CALD", "Indigenous", "Caucasian", "Not recorded"))) %>%
  group_by(Care_status, Race.nat) %>%
  summarise(n_indiv = n_distinct(MP_id)) %>%
  ggplot(  etc.  )
```

## Plotting two data sets of different length on one plot

https://gist.github.com/romunov/47f51ccdfe2362c66e60743849fde6b0

```{r eval = F}

xy1 <- data.frame(x = rnorm(10), y = rnorm(10))
xy2 <- data.frame(x = rnorm(5), y = rnorm(5))

library(ggplot2)

ggplot() +
  theme_bw() +
  geom_point(data = xy1, aes(x = x, y = y)) +
  geom_point(data = xy2, aes(x = x, y = y), color = "red")

ggplot(xy1, aes(x = x, y = y)) +
  theme_bw() +
  geom_point() + # values mapped to aes() in ggplot() call
  geom_point(data = xy2, color = "red") # values maped to the same aes() call as above

ggplot(xy1, aes(x = x, y = y)) +
  theme_bw() +
  geom_point() + # values mapped to aes() in ggplot() call
  geom_point(data = xy2, aes(x = x, y = y)) # explicit mapping, can be different for this geom


##  PGS' extra

ggplot(xy1, aes(x = x, y = y)) +
  theme_bw() +
  geom_point() + # values mapped to aes() in ggplot() call
  geom_point(
    data = xy2, aes(x = x, y = y), colour = "green") # explicit mapping, can be different for this geom

```

